---
title: 'Case Study 3: Patterns in DNA'
output:
  pdf_document: default
  html_notebook: default
---

```{r}
# CMD + SHIFT + K
# CMD + OPTION + I
# CMD + SHIFT + ENTER
```

```{r}
library(lattice)
library(purrr)
```

## 1 Locations
### 1.1 Hello
```{r}
DNA.len <- 229354

# Import the data
data <- read.table("data.txt", header=TRUE)

locs = data$location
locs.len = length(obs_locs)
locs.min = min(obs_locs)
locs.max = max(obs_locs)

# Build the generated palindorme locations
r_locs <- round(runif(locs.len, 0, DNA.len))
```

```{r}
# Plotting the observed locations
stripplot(locs, main='Location of Observed Palindromes', xlab='Locations', pch=16, cex=0.25) #one-dimensional scatter plot
stripplot(r_locs, main='Location of Generated Palindromes', xlab='Locations', pch=16, cex=0.25, col='red') #one-dimensional scatter plot
hist(locs, breaks = 30, xaxt='n', col=rgb(0,0,1,0.7), xaxt='n')
hist(r_locs, breaks = 30, col=rgb(1,0,0,0.7), xaxt='n', add=T)
axis(1, at=seq(0, DNA.len + 10000, 10000))
```

```{r}
# Split the data into intervals
indx = 1
intervals.size <- 2050
intervals <- split(locs, cut(locs, seq(locs.min, locs.max, by=intervals.size), 
                             include.lowest=TRUE, drop=FALSE))
intervals.num <- length(intervals)

# Count the total number of palindromes in our intervals
intervals.tot_pals = 0
for (intv in intervals) {
  intervals.tot_pals = intervals.tot_pals + length(intv)
}

lambda.est = intervals.tot_pals/intervals.num

# Print metrics
paste("Interval Size:", intervals.size)
paste("Number of intervals:", intervals.num)
paste("Total Palindromes (within intervals):", intervals.tot_pals)
paste("Lambda Est:", lambda.est)
```

```{r}
k_plus = 6

# Calculated the expected palindrome counts for {0, 1, ... 8, 9+}
pal.counts_est <- numeric(k_plus)

for (i in 0:k_plus) {
  if (i != k_plus) {
    pal.counts_est[i + 1] = (ppois(i, lambda.est, lower=T) - 
      ppois(i - 1, lambda.est, lower=T)) * intervals.num
  } else {
    pal.counts_est[i + 1] = ppois(k_plus - 1, lambda.est, lower=F) * intervals.num
  }
}

paste("Pal Count Estimates:", toString(pal.counts_est))
```

```{r}
# Get raw counts for each k palindromes
pal.counts <- integer(intervals.num)
i = 1
for (p in intervals) {
  pal.counts[i] = length(p)
  i = i + 1
}

title = paste('Observed Palindrome Counts Across', intervals.len, 'Intervals of Length', intervals.size)
barplot(pal.counts, col=rgb(0,0,1, 0.6), main=title, xlab=xlab)
```

```{r}
# Get raw counts for each k palindromes
r_locs <- round(runif(locs.len, 0, DNA.len))
r_intervals <- split(r_locs, cut(r_locs, seq(min(r_locs), max(r_locs), by=intervals.size), 
                             include.lowest=TRUE, drop=FALSE))

r_pal.counts <- integer(intervals.num)
i = 1
for (p in r_intervals) {
  r_pal.counts[i] = length(p)
  i = i + 1
}

title = paste('Generated Palindrome Counts Across', intervals.len, 'Intervals of Length', intervals.size)
barplot(r_pal.counts, col=rgb(1,0,0, 0.6), main=title, xlab=xlab)
```


```{r}
pal.counter <- integer(k_plus + 1)

# Build observed palindrome counts for contigency table
for (count in pal.counts) {
  if ( count >= k_plus ) {
    pal.counter[k_plus + 1] = pal.counter[k_plus + 1] + 1
  } else {
    pal.counter[count + 1] = pal.counter[count + 1] + 1
  }
}

print("Observed Palindrome Counts")
print(pal.counter)
```

```{r}
print("Contigency Table")
for (i in seq_along(pal.counter) ) {
  plus = " "
  if(i == k_plus + 1) {
    plus = "+"
  }
  str <- paste(toString(i - 1), plus, ": ", toString(pal.counter[i]), 
               "    ,", toString(pal.counts_est[i]))
  print(str)
}
```

```{r}
# Manually combine rows
n <- k_plus - 1 # Number of rows we'll be sizing down to
contigency_obs = numeric(n)
contigency_exp = numeric(n)
num_comb <- 1
for (i in 1:length(pal.counter)) {
  if (i <= num_comb) {
    contigency_obs[1] <- contigency_obs[1] + pal.counter[i]
    contigency_exp[1] <- contigency_exp[1] + pal.counts_est[i]
  } else {
    contigency_obs[i - num_comb + 1] <- pal.counter[i]
    contigency_exp[i - num_comb + 1] <- pal.counts_est[i]
  }
}

# Truncated Contigency Table 
pal_cont_table <- data.frame("Obs Pal Counts"=contigency_obs, "Exp Pal Counts"=contigency_exp)
pal_cont_table
```


## 1.3 Chi Squarted Function
```{r}
# Compute the chi-squared
#
# Args:
#     obs: List of observed values
#     exp: List of expected values
chi_sqrd <- function(obs, exp, k) {
  n <- length(obs)
  x <- 0
  for(i in 1:length(obs)) {
    x <- x + ((obs[i] - exp[i])^2)/(exp[i])
  }
  #x <- sum( ((obs - exp)^2)/exp)
  
  print(n - k)
  return (c(x, pchisq(x, n - k, lower=F)))
}
```

```{r}
sample_obs <- c(7, 8, 10, 9, 8, 5, 4, 6)
sample_exp <- c(6.4, 7.5, 9.7, 10, 8.6, 6.3, 4.1, 4.5)
print(chi_sqrd(contigency_obs, contigency_exp, 2))
```

### 1.6 Residuals of Palindrome Counts
```{r}
Residuals <- (contigency_obs - contigency_exp) / sqrt(contigency_exp)
plot(Residuals, type = 'h', ylab = "standardized residuals", xlab = "interval index")
```


```{r}
r_locs = round(runif(num_locs, min_loc, max_loc))
```

## Classifying Regions According to Counts
```{r}
tmp_i_size <- c(200, 500, 1000, 2050)
r_locs <- round(runif(locs.len, 0, DNA.len))
tmp.counts <- integer()

for (i_size in tmp_i_size) {
  tmp.counts <- integer()
  tmp <- split(locs, cut(locs, seq(min(locs), max(locs), by=i_size), 
                               include.lowest=TRUE, drop=FALSE))
  
  tmp.len <- length(tmp)
  tmp.counts <- integer(14)
  
  i = 1
  for (p in tmp) {
    if(length(p) >= 14 ){
      tmp.counts[14] = tmp.counts[14] + 1
    } else {
      tmp.counts[length(p)] = tmp.counts[length(p)] + 1
    }
    i = i + 1
  }
  
  title = paste('Observed Palindrome Counts Across', tmp.len, 'Intervals of Length', i_size)
  bp <- barplot(tmp.counts, col=rgb(0,0,1, 0.6), main=title, ylim=c(0,max(tmp.counts) + 10), xaxt='n', xlab="Palindrome Counts", ylab="Freq")
  
  labels <- (1:14)
  labels[14] <- "14+"
  text(x=bp[,1], y=-1, adj=c(1, 1), labels, cex=0.8, srt=45, xpd=TRUE)
}
```


## 2 Spacing Between Palindromes
### 2.1 Compute and Generate Spacings
```{r}
# Calculate observed spacings
pair_spacings <- diff(locs)
trip_spacings <- numeric(length(pair_spacings) - 1)
for(i in 1:(length(pair_spacings) - 1)) {
  trip_spacings[i] <- pair_spacings[i] + pair_spacings[i+1]
}

# Compute lambda estimate
exp_lambda.est <- 1 / mean(pair_spacings)

# Generate spacings
r_pair_spacings <- rexp(length(pair_spacings), exp_lambda.est)
r_trip_spacings <- rgamma(length(trip_spacings), 2, exp_lambda.est)
```

### 2.2 Plot Spacings Between Palindromes
```{r}
# Histogram of Observed Spacings
hist(pair_spacings, breaks = 30, xaxt='n',
     main="Histogram of Observed Spacings Between Pairs and Triplets", 
     col=rgb(0,0,1,0.7), xlab ="Number of Basepairs Apart")
hist(trip_spacings, col=rgb(1,0,0,0.7),  xaxt='n', add=T)
axis(1, seq(0, max(trip_spacings), 200))
legend("topright", c("Pairs", "Triplets"), col=c(rgb(0,0,1,0.7), rgb(1,0,0,0.7)), lwd=8)

# Histogram of Generated Spacings
hist(r_pair_spacings, breaks = 30, xaxt='n', 
     main="Histogram of Generated Spacings Between Pairs and Triplets", 
     col=rgb(0,0,1,0.7), xlab ="Number of Basepairs Apart", xaxt='n')
hist(r_trip_spacings, col=rgb(1,0,0,0.7), add=T)
axis(1, seq(0, max(r_pair_spacings), 200))
legend("topright", c("Pairs", "Triplets"), col=c(rgb(0,0,1,0.7), rgb(1,0,0,0.7)), lwd=8)

# Histogram Comparison of Observed and Generated Pair Spacings
hist(pair_spacings, breaks = 30, xaxt='n',
     main="Comparison of Observed and Generated Pair Spacings", 
     col=rgb(0,0,1,0.7), xlab ="Number of Basepairs Apart")
hist(r_pair_spacings, breaks = 30, xaxt='n', 
     col=rgb(1,0,0,0.7), add=T)
axis(1, seq(0, max(c(pair_spacings, r_pair_spacings)), 200))
legend("topright", c("Observed", "Generated"), col=c(rgb(0,0,1,0.7), rgb(1,0,0,0.7)), lwd=8)

# Histogram Comparison of Observed and Generated Triplet Spacings
hist(trip_spacings, main="Comparison of Observed and Generated Triplet Spacings", 
     col=rgb(0,0,1,0.7), xlab ="Number of Basepairs Apart", yaxt='n',  xaxt='n', ylim=c(0, 75))
hist(r_trip_spacings, xaxt='n', col=rgb(1,0,0,0.7), xlab ="Number of Basepairs Apart", yaxt='n', add=T )
axis(1, seq(0, max(c(trip_spacings, r_trip_spacings)) + 200, 200))
axis(2, seq(0, 70, 10))

legend("topright", c("Observed", "Generated"), col=c(rgb(0,0,1,0.7), rgb(1,0,0,0.7)), lwd=8)
```


### 2.3 Create Contigency Table for Spacings
```{r}
sizes <- 75
len <- 750
num_len_above <- sum((pair_spacings > len))

a <- split(pair_spacings, cut(pair_spacings, seq(0, len, by=sizes), include.lowest=TRUE, drop=FALSE))
a.count <- c(unlist(map(a, length), use.names=FALSE), num_len_above)

exp_spaces <- numeric(length(a.count))

j = 1
for(i in seq(0, len, by=sizes)) {
  if (i != len) {
    exp_spaces[j] <- pexp(i + sizes, exp_lambda.est, lower=T) - pexp(i, exp_lambda.est, lower=T)
  } else {
    exp_spaces[j] <- pexp(len, exp_lambda.est, lower=F)
  }
  j = j + 1
}
exp_spaces = exp_spaces*length(pair_spacings)
```

```{r}
space_cont_table <- data.frame("Bucket"=seq(0, len, by=sizes), 
                               "Obs counts"=a.count, 
                               "Exp Counts"=exp_spaces)
space_cont_table
chi_sqrd(a.count, exp_spaces, 2)
```
```{r}
Residuals <- (a.count - exp_spaces) / sqrt(exp_spaces)
plot(Residuals, type = 'h', main="Residuals for Spaces Between Palindrome Pairs", ylab = "standardized residuals", xlab = "interval index")
```

```{r}
stripplot(pair_spacings)
```

### 2.4 Create Contigency Table for Triplet Spacings
```{r}
sizes <- 150
len <- 1500
num_len_above <- sum((trip_spacings > len))

b <- split(trip_spacings, cut(trip_spacings, seq(0, len, by=sizes), include.lowest=TRUE, drop=FALSE))
b.count <- c(unlist(map(b, length), use.names=FALSE), num_len_above)

exp_trip_spaces <- numeric(length(b.count))
j = 1
for(i in seq(0, len, by=sizes)) {
  if (i != len) {
    exp_trip_spaces[j] <- pgamma(i + sizes, 2, exp_lambda.est, lower=T) - pgamma(i,2, exp_lambda.est, lower=T)
  } else {
    exp_trip_spaces[j] <- pgamma(len, 2, exp_lambda.est, lower=F)
  }
  j = j + 1
}

exp_trip_spaces = exp_trip_spaces * length(trip_spacings)
```

```{r}
space_cont_table <- data.frame("Bucket"=seq(0, len, by=sizes), 
                               "Obs counts"=b.count, 
                               "Exp Counts"=exp_trip_spaces)
space_cont_table
print(chi_sqrd(b.count, (exp_trip_spaces), 2))
```

```{r}
Residuals <- (a.count - exp_spaces) / sqrt(exp_spaces)
plot(Residuals, type = 'h', main="Residuals for Spaces Between Palindrome Triplets", ylab = "standardized residuals", xlab = "interval index")
```



## 3 Locations and Uniform Distribution
### 3.1 Compute Observed and Generated Count of Palindromes Across Buckets
```{r}
num_buckets <- 8
r_locs <- round(runif(locs.len, 0, DNA.len))

split_locs <- split(locs, cut(locs, seq(0, DNA.len, by=(DNA.len/num_buckets)), include.lowest=TRUE, drop=FALSE))
split_locs.count <- unlist(map(split_locs, length), use.names=FALSE)

r_split_locs <- split(r_locs, cut(r_locs, seq(0, DNA.len, by=(DNA.len/num_buckets)), include.lowest=TRUE, drop=FALSE))
r_split_locs.count <- unlist(map(r_split_locs, length), use.names=FALSE)

barplot(split_locs.count,  main=paste("Palindrome Counts Across",num_buckets, "Intervals"), col=rgb(0,0,1,0.7))
barplot(r_split_locs.count, col=rgb(1,0,0,0.7), add=T)
legend("topleft", c("Observed", "Generated"), col=c(rgb(0,0,1,0.7), rgb(1,0,0,0.7)), lwd=2)
```

### 3.2 Contigency Table of Palindrome Locations
```{r}
exp_counts <- replicate(num_buckets, locs.len/num_buckets)
obs_counts <- split_locs.count

contigency_table <- data.frame("Bucket #"=1:num_buckets, "Obs counts"=obs_counts, "Exp Counts"=exp_counts)
contigency_table
```

### 3.3 P-Value of Chi-Squared Test for Palindrome Location Counts
```{r}
print(chi_sqrd(obs_counts, exp_counts, 2))
```

### 3.4 Residuals of Palindrome Location Counts
```{r}
Residuals <- (obs_counts - exp_counts) / sqrt(exp_counts)
plot(Residuals, type = 'h', main="Residuals for Palindrome Location Counts", ylab = "standardized residuals", xlab = "interval index")
```

